### 2020-05-12 serial.tex
minor tweaks

Generation and serial intervals are under-appreciated yet important quantities for charaterizing the outset of an outbreak.
Even though their definitions (and differences) were clearly laid out over a decade ago, there is still a need for clear conceptual and theoretical understanding of their roles in shaping the epidemiological dynamics.
Part of the confusion can be attributed to an apparent paradox.
By definition, both generation and serial intervals describe how infection spreads between individuals --- thus, one would expect them both to link between the epidemic growth rate $r$ and the reproduction number \RR.
However, current theory does not explain how two different distributions can give the same link between $r$ and \RR --- some studies have further demonstrated that generation and serial intervals give different $r$--\RR link. \jd{Revisit}
We provide an answer to the paradox by showing that \emph{forward-looking} serial intervals, which correctly link the two quantities, are not the same as ``intrinsic'' serial intervals, but instead change with $r$. 
We provide a heuristic way of addressing potential biases that can arise from not accounting for changes in these references and apply our framework to serial intervals of COVID-19.
This study demonstrates the importance of early epidemiological investigation through contact tracing for characterizing initial serial intervals and provides rationale for reassessing serial intervals, and \RR\ estimates, for COVID-19.

@@

The reproduction number \RR is one of the most important characteristics of an emerging epidemic such as novel coronavirus disease (COVID-19) \citep{majumder2020early}.
The reproduction number is defined as the average number of secondary cases caused by a primary case;
the reproduction number in a fully susceptible population --- referred to as the basic reproduction number \Ro --- allows us to predict the extent to which a disease will spread in the population and the amount of intervention necessary to eliminate disease \citep{anderson1991infectious}.
Near the beginning of an outbreak, \RR is often estimated from the observed exponential growth rate using generation- and serial-interval distributions (e.g., \cite{du2020serial, jung2020real, li2020early, zhao2020preliminary}).

@@

At the individual level, we can define the expected time distribution between a primary and a secondary event --- we refer to this distribution as the intrinsic distribution.
For example, the intrinsic incubation period distribution describes the expected time from infection to symptom onset of an individual.
Likewise, the intrinsic generation-interval distribution describes the expected time from infection of a primary case to infection of a secondary case in a fully susceptible population; 
since this is a expected time distribution, it does not take into account dynamical factors such as susceptible depletion or intervention efforts.
However, the intrinsic time distributions are not always equivalent to realized time distributions at the population level.

@@

At the individual level, both forward and backward \emph{perspectives}
must yield identical results (e.g., the length of the incubation
period of a given individual is the same whether measured forward from
the time of infection or backward from the time of symptom onset).
Consequently, if we restrict attention to individuals with a given
incubation period, say $\tau$, and we count how many became infected at time $\psymp$
or how many first showed symptoms at time $\ssymp=\psymp+\tau$ then we must obtain
the same number. In general, no matter how delays are distributed, if
$\mathcal P$ and $\mathcal S$ represent the sizes of primary and
secondary cohorts then we must have
\begin{equation}
\psize(\psymp) f_\psymp(\tau) = \ssize(\ssymp) b_\ssymp(\tau) \,.
\label{eq:match}
\end{equation}
Substituting $\psymp=\ssymp-\tau$, it follows that
\begin{equation}
b_\ssymp(\tau)= \frac{\psize(\ssymp-\tau) f_{\ssymp-\tau} (\tau)}{\ssize(\ssymp)}
\label{eq:backward}
\end{equation}
If we are thinking of incubation periods, then the left hand side of
this equation is the probability density that an individual
who became symptomatic at time $\ssymp$ had an incubation period of
length $\tau$. From the right hand side, we see that this probability
density depends on how many individuals became infected at time
$\ssymp-\tau$ [$\psize(\ssymp-\tau)$], how many individuals became
symptomatic at time $\ssymp$ [$\ssize(\ssymp)$], and the probability density
that someone who was infected at time $\ssymp-\tau$ became symptomatic at
time $\ssymp$ [$f_{\ssymp-\tau}(\tau)$]. In general, Eq.~\eqref{eq:backward}
shows that the the backward delay distribution at reference time $\ssymp$
[$b_\ssymp(\tau)$] depends not only on the secondary cohort size at time
$\ssymp$ [$\ssize(\ssymp)$], but also on the primary cohort size at the
earlier time $\ssymp-\tau$ [$\psize(\ssymp-\tau)$] and the forward delay
distribution at that earlier time [$f_{\ssymp-\tau}(\tau)$].

@@

To calculate realized serial-interval distributions, we begin by modeling $\total(\psymp,\ssymp)$: the total number of serial intervals that start (when infectors develop symptoms) at time $\psymp$ and end (when infectees develop symptoms) at time $\ssymp$.
The number of serial intervals between time $\psymp$ and $\ssymp$ given that the infectors became infected at time $\pinf$ and the infectees became infected at time $\sinf$ depends on the amount of infection that occur between time $\pinf$ and $\sinf$ as well as the densities of forward incubation period between $\pinf$ and $\psymp$ (incubation periods of infectors) as well as $\sinf$ and $\ssymp$ (incubation periods of infectees):
\begin{equation}
\underbrace{i(\pinf)}_{\text{incidence}} 
\times 
\underbrace{\Rc (\pinf)}_{\substack{\text{case} \\ \text{reproduction} \\ \text{number}}} 
\times 
\underbrace{h_{\pinf}(\psymp-\pinf, \sinf - \pinf)}_{\substack{\text{joint density of} \\ \text{the forward incubation} \\ \text{period } p-\pinf \text{ and the forward} \\ \text{generation interval } \sinf-\pinf\\ \text{(of infectors)}}}
\times
\underbrace{k_{\sinf}(\ssymp - \sinf)}_{\substack{\text{marginal density of} \\ \text{the forward incubation} \\ \text{period } \ssymp-\sinf \\ \text{(of infectees)}}},
\end{equation}
where the case reproduction number $\Rc (\pinf)$ is defined as the average number of secondary cases caused by a primary case infected at time $\pinf$ over the course of their infection \citep{fraser2007estimating}.
We note that it is necessary to describe the forward incubation periods and the forward generation intervals using a joint probability distribution because onset of symptoms and transmission potential jointly depend on the life history of a disease;
for example, if an infected individual can only transmit the disease after symptom onset, the forward generation interval will necessarily be longer than the forward incubation period.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=0.95\textwidth]{forward.pdf}
\caption{
\textbf{Epidemiological dynamics and changes in mean forward and backward delay distributions.}
(A) Cumulative incidence over time. \jd{Would it be easier for people to think about dynamics if we show incidence instead?}
(B--D) Changes in the mean forward incubation period, generation interval, and serial interval.
(E--G) Changes in the mean backward incubation period, generation interval, and serial interval.
Black lines represent the results of a deterministic simulation.
Orange lines and points represent the average of 10 stochastic simulations;
cumulative incidence from the stochastic simulations in panel A are shown without averaging.
\jd{I don't see orange lines, and the location of the dashed line is not explained. I guess you use the forward value at time 0? Would the intrinsic value be better?}
Forward incubation periods and intrinsic generation-intervals are assumed to be independent of each other.
See Table 1 for parameter values.
}
\label{fig:epi}
\end{center}
\end{figure}

>>>>

Generation and serial intervals are under-appreciated yet important quantities for charaterizing the outset of an outbreak.
Even though their definitions (and differences) were clearly laid out over a decade ago, there is still a need for clear conceptual and theoretical understanding of their roles in shaping the epidemiological dynamics.
Part of the confusion can be attributed to an apparent paradox.
By definition, both generation and serial intervals describe how infection spreads between individuals --- thus, one would expect them both to link between the epidemic growth rate $r$ and the reproduction number \RR.
However, current theory does not explain how two different distributions can estimate the same \RR from $r$ --- some studies have further demonstrated that generation and serial intervals give different estimates of \RR.
We provide an answer to the paradox by showing that \emph{forward-looking} serial intervals, which correctly estimates \RR from $r$, are not the same as ``intrinsic'' serial intervals, but instead change with $r$. 
We provide a heuristic way of addressing potential biases that can arise from not accounting for changes in these references and apply our framework to serial intervals of COVID-19.
This study demonstrates the importance of early epidemiological investigation through contact tracing for characterizing initial serial intervals and provides rationale for reassessing serial intervals, and \RR\ estimates, for COVID-19.

@@

The reproduction number \RR is one of the most important characteristics of an emerging epidemic such as Coronavirus Disease 2019 (COVID-19) \citep{majumder2020early}.
The reproduction number is defined as the average number of secondary cases caused by a primary case;
the reproduction number in a fully susceptible population --- referred to as the basic reproduction number \Ro --- allows us to predict the extent to which a disease will spread in the population and the amount of intervention necessary to eliminate disease \citep{anderson1991infectious}.
Near the beginning of an outbreak, \RR is often estimated from the observed exponential growth rate using generation- and serial-interval distributions (e.g., \cite{du2020serial, jung2020real, li2020early, zhao2020preliminary}).

@@

At the individual level, we can define the expected time distribution between a primary and a secondary event for an average infected individual --- we refer to this distribution as the intrinsic distribution.
For example, the intrinsic incubation period distribution describes the expected time distribution from infection to symptom onset of an average infected individual.
Likewise, the intrinsic generation-interval distribution describes the expected time distribution of infectious contacts made by an average infected individual.
However, the intrinsic time distributions are not always equivalent to the corresponding realized time distributions at the population level (i.e., distribution of time between primary and secondary events that occur during an epidemic).
For example, an infectious contact may not result in infection if the contacted individual is not susceptible;
therefore, the realized generation-interval distribution can be different from the intrinsic generation-interval distribution.

@@

At the individual level, both forward and backward \emph{perspectives}
must yield identical measurement (e.g., the length of the incubation
period of a given individual is the same whether measured forward from
the time of infection or backward from the time of symptom onset).
Consequently, if we restrict attention to individuals with a given
incubation period, say $\tau$, and we count how many became infected at time $\psymp$
or how many first showed symptoms at time $\ssymp=\psymp+\tau$ then we must obtain
the same number. In general, no matter how delays are distributed, if
$\mathcal P$ and $\mathcal S$ represent the sizes of primary and
secondary cohorts then we must have
\begin{equation}
\psize(\psymp) f_\psymp(\tau) = \ssize(\ssymp) b_\ssymp(\tau) \,.
\label{eq:match}
\end{equation}
Substituting $\psymp=\ssymp-\tau$, it follows that
\begin{equation}
b_\ssymp(\tau)= \frac{\psize(\ssymp-\tau) f_{\ssymp-\tau} (\tau)}{\ssize(\ssymp)}
\label{eq:backward}
\end{equation}
If we are thinking of realized incubation periods, then the left hand side of
this equation is the probability density that an individual
who became symptomatic at time $\ssymp$ had an incubation period of
length $\tau$. From the right hand side, we see that this probability
density depends on how many individuals became infected at time
$\ssymp-\tau$ [$\psize(\ssymp-\tau)$], how many individuals became
symptomatic at time $\ssymp$ [$\ssize(\ssymp)$], and the probability density
that someone who was infected at time $\ssymp-\tau$ became symptomatic at
time $\ssymp$ [$f_{\ssymp-\tau}(\tau)$]. In general, Eq.~\eqref{eq:backward}
shows that the the backward delay distribution at reference time $\ssymp$
[$b_\ssymp(\tau)$] depends not only on the secondary cohort size at time
$\ssymp$ [$\ssize(\ssymp)$], but also on the primary cohort size at the
earlier time $\ssymp-\tau$ [$\psize(\ssymp-\tau)$] and the forward delay
distribution at that earlier time [$f_{\ssymp-\tau}(\tau)$].

@@

To calculate realized serial-interval distributions, we begin by modeling $\total(\psymp,\ssymp)$: the total number of serial intervals that start (when infectors develop symptoms) at time $\psymp$ and end (when infectees develop symptoms) at time $\ssymp$.
The number of serial intervals between time $\psymp$ and $\ssymp$ given that the infectors became infected at time $\pinf$ and the infectees became infected at time $\sinf$ depends on the amount of infection that occur between time $\pinf$ and $\sinf$ as well as the densities of forward incubation period between $\pinf$ and $\psymp$ (incubation periods of infectors) as well as $\sinf$ and $\ssymp$ (incubation periods of infectees):
\begin{equation}
\underbrace{\Rc (\pinf)}_{\substack{\text{case} \\ \text{reproduction} \\ \text{number}}} 
\times 
\underbrace{i(\pinf)}_{\text{incidence}} 
\times 
\underbrace{h_{\pinf}(\psymp-\pinf, \sinf - \pinf)}_{\substack{\text{joint density of} \\ \text{the forward incubation} \\ \text{period } p-\pinf \text{ and the forward} \\ \text{generation interval } \sinf-\pinf\\ \text{(of infectors)}}}
\times
\underbrace{k_{\sinf}(\ssymp - \sinf)}_{\substack{\text{marginal density of} \\ \text{the forward incubation} \\ \text{period } \ssymp-\sinf \\ \text{(of infectees)}}},
\end{equation}
where the case reproduction number $\Rc (\pinf)$ is defined as the average number of secondary cases caused by a primary case infected at time $\pinf$ over the course of their infection \citep{fraser2007estimating}.
We note that it is necessary to describe the forward incubation periods and the forward generation intervals using a joint probability distribution because onset of symptoms and transmission potential jointly depend on the life history of a disease;
for example, if an infected individual can only transmit the disease after symptom onset, the forward generation interval will necessarily be longer than the forward incubation period.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=0.95\textwidth]{forward.pdf}
\caption{
\textbf{Epidemiological dynamics and changes in mean forward and backward delay distributions.}
(A) Daily incidence over time.
(B--D) Changes in the mean forward incubation period, generation interval, and serial interval.
(E--G) Changes in the mean backward incubation period, generation interval, and serial interval.
Black lines represent the results of a deterministic simulation.
Orange lines (A) represent the results of 10 stochastic simulations.
Orange points (B-G) represent the average of 10 stochastic simulations.
\jd{I don't see orange lines, and the location of the dashed line is not explained. I guess you use the forward value at time 0? Would the intrinsic value be better?}
Intrinsic incubation periods and intrinsic generation-intervals are assumed to be independent of each other.
See Table 1 for parameter values.
}
\label{fig:epi}
\end{center}
\end{figure}
